
---
output: github_document
---

---
title: "REPORT: ANALYSIS OF THE OTTER GROOMING PATTERNS"
output: word_document
date: '2022-05-12'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(treemap)

```

## Introduction

### Bacground Information
The following project was used to analyze the grooming patterns of Nothern America’s River Otters. Before doing any analysis on the dataset, it was a important to have some background information on the river residing animals. River otters can be diurnal or nocturnal, with most being more active at night. 
Giant otters are only active during the day. Clawless otters are primarily nocturnal, though some individuals may be active during the day in remote areas where humans are not present. Sea otters are mostly active during the day. Feeding and grooming are the primary daily activities, which are interspersed with rest periods. Note that the claws aspects and sleeping patterns of otters are not explored for this project.
Grooming
To keep their fur insulating, all otters must groom it constantly. River otters spend a significant amount of time grooming, and many species have designated areas on land for drying and grooming their fur. The majority of them dry themselves vigorously by rolling on the ground or rubbing against logs or vegetation.
Researchers discovered that sea otters spend at least 11% to 48% of their day grooming. They comb their fur and remove debris with their paws and claws. They can also aerate their fur by blowing air into it and pounding the water with their feet to create foam. Because of its flexible body and loose-fitting skin, an otter can reach every part of its fur.

### The Dataset
The dataset had the following features:

**Group-This was a group of otters that was watched at one time. For instance group A had one female and 3 males**
**Season -There were only two categories; whether it was the breeding season or not**
**Time - The time in minutes the group was watched**
**Groomer - The otter that did the grooming. They were named by gender and a number was asssigned**
**Groomee - The otter that was groomed. The same naming convention was as the groomers**
**Frequency – The number of instances of grooming that occurred.**
## objectives
The project had the following objectives:
1.	Determine whether animals in each group groomed equally
2.	Identify any preferences in terms of grooming in each group.
3.	Determine if males groomed more than females
4.	Mark any differences in the grooming rates in each season
## Tools Used
The analysis of the otter dataset was done in R and relied heavily on the Tidyverse package which included the ggplot plotting libarary that was used in the visualization that are featured in this report.

## Data Engineering
Before the dataset could be summarized into plots, it was first mutated into a tibble to allow for better handling and summarizing of the data. 
The data was provided in the format of an R datafile.
The first action that was done was to find the genders of the groomers and groomees. `Case_when` method was used to determine the genders of each otter since the first letter indicated the otter’s gender.

```{r}
load("classdata.RData")
otter = as.data.frame(otter)
head(otter)

otter = as_tibble(otter)
head(otter)
# Adding column based on other column:
otter_df = otter %>%
  mutate(groomer_gender = case_when(
    startsWith(otter$groomer, "F") ~ "FEMALE",
    startsWith(otter$groomer, "M") ~ "MALE"
  ))%>%
  mutate(groomee_gender = case_when(
    startsWith(otter$groomee, "F") ~ "FEMALE",
    startsWith(otter$groomee, "M") ~ "MALE"
  ))
otter_df %>% 
  head() %>%
  kable()
```
Before proceeding the Groomer and Gromee otters were paired together to form mates column:




```{r}
otter_df = otter_df  %>% 
  mutate(otter_df, mates = paste(otter_df$groomer, otter_df$groomee))

otter_df %>% 
  head() %>%
  kable()
```


```{r}
otter_df = otter_df %>%
  mutate(mating_type = case_when(
    startsWith(otter_df$groomer_gender, "F")& startsWith(otter_df$groomee_gender, "M") ~ "HETEROSEXUAL",
    startsWith(otter_df$groomer_gender, "M")& startsWith(otter_df$groomee_gender, "F") ~ "HETEROSEXUAL",
    startsWith(otter_df$groomer_gender, "F")& startsWith(otter_df$groomee_gender, "F") ~ "HOMOSEXUAL",
    startsWith(otter_df$groomer_gender, "M")& startsWith(otter_df$groomee_gender, "M") ~ "HOMOSEXUAL",
  ))
otter_df %>% 
  head() %>%
  kable()
```

## Analysis:
### Grooming Patterns of Each Group
In the following two plots the first two objectives are met, that is:
•	Determine whether animals in each group groomed equally
•	Identify any preferences in terms of grooming in each group.
The first objective wa to determine if the animals in each group groomed equally. For instance, in Group A there was one female and three males. Did the female groom equally with all the females? 
It is important to note that for each group the female partner had multiple partners.
The female otter groomed with all the males in the group, however, the highest number of grouping was done with M2. The least number of grouping was done between the males. The otters thefore do not groom equally between each other.

```{r, echo=FALSE}
# Build Dataset
group <- paste("Group" , otter_df$group, sep="-")
subgroup <-otter_df$mates
value <- otter_df$frequency
data <- data.frame(group,subgroup,value)

# Custom labels:
treemap(data, index=c("group","subgroup"),     vSize="value", type="index",
        
        fontsize.labels=c(15,12),                # size of labels. Give the size per level of aggregation: size for group, size for subgroup, sub-subgroups...
        fontcolor.labels=c("white","black"),    # Color of labels
        fontface.labels=c(2,1),                  # Font of labels: 1,2,3,4 for normal, bold, italic, bold-italic...            # Background color of labels
        align.labels=list(
          c("center", "center"), 
          c("right", "bottom")
        ),                                   # Where to place labels in the rectangle?
        overlap.labels=0.5,                      # number between 0 and 1 that determines the tolerance of the overlap between labels. 0 means that labels of lower levels are not printed if higher level labels overlap, 1  means that labels are always printed. In-between values, for instance the default value .5, means that lower level labels are printed if other labels do not overlap with more than .5  times their area size.
        inflate.labels=F,                        # If true, labels are bigger when rectangle is bigger.
        title = "Tree Map Showing Mating Frequency Group",
        border.col=c("black","white"),             # Color of borders of groups, of subgroups, of subsubgroups ....
        border.lwds=c(5,2),                            # How you color the treemap. type help(treemap) for more info
        palette = "Set2",                        # Select your color palette from the RColorBrewer presets or make your own.
        fontsize.title=12,
)
```

We also note that the frequency of grooming has reduces significantly when it is not the breeding season. Although the pattern of which otter grooms one with the other remains the same. From the plot below we note that the otters maintain their preference in the partners they groom with.

```{r, echo=FALSE}
otter_df %>% # data # manipulation
  ggplot()+
  geom_bar(aes(x=mates, y = frequency, fill = group), # separate by category
           stat = "summary", 
           position="dodge")+ coord_flip()+ # make it side-by-side
         theme_light()+facet_wrap(~season)+ggtitle("Mating Frequency in Each Season")
```


### Grooming Patterns For Each Gender:

```{r, echo=FALSE}
otter_df %>%
  group_by(groomee_gender)%>%
  summarize(n=n())%>%
  mutate(prop=n/sum(n)) %>%
  kable()
```

Females were observed to groomers 38.6% percent of the times.However, it is important to note that the number of males in this dataset is higher than that of females. The proportion does not change whether its the breeding season or not. The bar plot below for each group gives a more objective look at groomers for each group in each season

```{r}
otter_df %>% # data # manipulation
  ggplot()+
  geom_bar(aes(x=season, y = frequency, 
               fill=groomer_gender), # separate by category
           stat = "summary", 
           position="dodge")+ # make it side-by-side
         theme_light()+facet_wrap(~group)+ggtitle("Grooming Patterns Group, Gender, and Season Wise Gender")

```


### Grooming Rates Per Season

```{r}
otter_df %>% # data # manipulation
  ggplot()+
  geom_bar(aes(x=season, y = frequency, 
               fill=groomee_gender), # separate by category
           stat = "summary", 
           position="dodge")+ # make it side-by-side
  theme_light()+ggtitle("Grooming Frequency For Each Season")
```

We note that the grooming frequency in the breeding season is higher, than when they are not breeding.

### Same Sex Grooming

Same Sex Grooming was also explored for each of the groups:

```{r}

# create a dataset
mating_type <- otter_df$mating_type
group <- otter_df$group
time <- otter_df$time
season = otter_df$season
data <- data.frame(mating_type,group,time, season)

# Grouped
ggplot(data, aes(fill=group, y=time, x=mating_type)) + 
  geom_bar(position="dodge", stat="identity") + facet_wrap(~season)+theme_light()+ggtitle("Observation Time for Each Group in each Season")

```

There were only two groups that displayed same sex grooming that is Group A and H
In terms of time, the time of observation of the groups was shorter in the breeding season except for group B.

```{r}
# create a dataset
mating_type <- otter_df$mating_type
group <- otter_df$group
frequency <- otter_df$frequency
season = otter_df$season
data <- data.frame(mating_type,group,frequency, season)

# Grouped
ggplot(data, aes(fill=group, y=frequency, x=mating_type)) + 
  geom_bar(position="dodge", stat="identity") + facet_wrap(~season)+theme_light()+ggtitle("Grooming Frequency For Each Season")
```

Same sex grooming was a bit higher in the Not breeding season while the opposite trend was observed in the heterosexual grooming(referring to grooming between different sexes), it was higher in the breeding season but dropped when it was not the breeding season.
## Conclusions

**1.	Note that for each group the female partner had multiple partners, therefore mated with the available otters.**
**2.	The least number of grouping was done between the males. The otters thefore do not groom equally between each other.**
**3.	We note that the grooming frequency in the breeding season is higher, than when they are not breeding.**
**4.	Same sex grooming was slightly higher during the non-breeding season, whereas heterosexual grooming (referring to grooming between different sexes) was higher during the breeding season but decreased during the non-breeding season.**


